---
title: Seasonal comparisons
author: "MW Beck"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document
---

```{r setup, echo = FALSE, message = F, warning = F, results = 'hide'}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.height = 8, fig.width = 9)

library(tidyverse)
library(tbeptools)
library(WRTDStidal)
library(lubridate)
library(mapview)
library(sf)
library(wqtrends)
library(patchwork)

data(rswqdat)
data(rsstatloc)
data(bswqdat)
data(ppseg)

source('R/funcs.R')

# # segments
# ppseg <- ppseg %>% 
#   rename(area = Name) %>% 
#   group_by(area) %>% 
#   summarise() %>% 
#   st_buffer(dist = 0.0001) %>% 
#   st_buffer(dist = -0.0001) %>% 
#   mutate(
#     area = factor(area)
#   ) %>% 
#   filter(area == 'Area')

var <- 'chla'
ylb <- expression(paste('Chl-',italic(a),' (',italic('\u03bc'),'g ',L^-1,')'))

# prep data, 2006 to present, all epc data
bsdat <- bswqdat %>% 
  filter(source == 'epchc') %>% 
  filter(date >= as.Date('2006-01-01'))

alldat <- rswqdat %>% 
  filter(source == 'epchc') %>% 
  filter(station %in% bsdat$station) %>% 
  bind_rows(bsdat) %>% 
  select(station, date, var, val) %>%
  # mutate(
  #   date = floor_date(date, unit = 'month')
  # ) %>% 
  filter(var %in% c(!!var, 'sal')) %>% 
  # group_by(station, date, var) %>% 
  # summarise(
  #   val = median(val, na.rm = T), 
  #   .groups = 'drop'
  # ) %>% 
  arrange(station, date) %>% 
  spread(var, val) %>%
  rename(var = !!var)
  # inner_join(rsstatloc, ., by = 'station') %>% 
  # select(-source_lng, -source, -comment)
```

# {.tabset}

## WRTDStidal approach {.tabset .tabset-pills}

These plots show model predicted chlorophyll as a function of time, season, and salinity using a WRTDS approach described [here](http://fawda123.github.io/WRTDStidal/articles/overview.html).  The results provide an estimate of chlorophyll trends at different conditional percentiles of the response, in this case, the median and 90th percentile (tau = 0.5, 0.9).  A "flow-normalized" trend independent of salinity is also provided.  The top plot shows predicted chlorophyll (lines) for the different percentiles overlaid on the observed values (points).  The middle plot shows flow-normalized chlorophyll (lines) against the predictions (points).  Importantly, predictions for 2021 show little difference between the flow-normalized and predicted results, suggesting little influence of freshwater inputs into the system.  The bottom two plots show the flow-normalized results by day of year, with separate lines by year.  Note the uptick in 2021 flow-normalized predictions (thicker line) of chlorophyll for some stations. 

### Station 21

```{r, results = 'hide'}
wrtdsfunc(alldat, 21, ylb)
```

### Station 22

```{r, results = 'hide'}
wrtdsfunc(alldat, 22, ylb)
```

### Station 90

```{r, results = 'hide'}
wrtdsfunc(alldat, 90, ylb)
```

## wqtrends approach {.tabset .tabset-pills}

These plots show model-predicted chlorophyll (as a function of time) using a GAMs approach described [here](https://tbep-tech.github.io/wqtrends/articles/introduction.html).  The basic idea is to model a signal of a long-term seasonal and annual trend from an observed time series and then extract a seasonal metric of interest from the modelled signal.  The seasonal metric applies to an a priori window of time (e.g., March, April, etc.) and includes statistical uncertainty that can be used for additional hypothesis testing.  The results do not include any flow-adjusted or hydrologic inputs. 

### Station 21

```{r, results = 'hide'}
gamfunc(alldat, 21, ylb)
```

### Station 22

```{r, results = 'hide'}
gamfunc(alldat, 22, ylb)
```

### Station 90

```{r, results = 'hide'}
gamfunc(alldat, 90, ylb)
```

## Observed and aggregated data

These plots show median plus 5th/95th percentiles of observed chlorophyll, total nitrogen, and total phosphorus for all stations in the immediate vicinity of Piney Point (Area 1 [here](https://tbep-tech.github.io/piney-point-analysis/trends#region-delineation)). Observations for 2021 are shown in red, whereas "baseline" observations are shown in black.  Baseline includes data for all long-term monitoring stations in Area 1 from 2006 to present.    

```{r, fig.height = 8, fig.width = 6}
area1 <- ppseg %>% 
  filter(Name == 'Area 1')
var <- c('tn', 'tp', 'chla', 'nh34', 'no23')

nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')

# prep data, 2006 to present, all epc data
bsdat <- bswqdat %>% 
  filter(date >= as.Date('2006-01-01'))

alldat <- rswqdat %>%
  bind_rows(bsdat) %>% 
  select(station, date, var, val) %>%
  filter(var %in% c(!!var)) %>% 
  arrange(station, date) %>% 
  group_by(station, date, var) %>% 
  summarise(
    val = median(val), 
    .groups = 'drop'
    ) %>% 
  spread(var, val) %>% 
  inner_join(rsstatloc, ., by = 'station') %>% 
  .[area1, ] %>% 
  st_set_geometry(NULL) %>% 
  filter(!station %in% nonbay) #%>% 
  # na.omit

toplo <- alldat %>% 
  gather('var', 'val', chla, tn, tp, no23, nh34) %>% 
  mutate(
    mo = month(date, abbr = T, label = T), 
    yday = yday(date), 
    yr = year(date), 
    typ = case_when(
      yr < 2021 ~ 'baseline (2006 - 2020)', 
      T ~ '2021'
    )
  ) %>% 
  group_by(typ, mo, var) %>% 
  summarise(
    medv = median(val, na.rm = T), 
    hiv = quantile(val, 0.05, na.rm = T), 
    lov = quantile(val, 0.95, na.rm = T), 
    .groups = 'drop'
  ) %>% 
  mutate(
    lbs = case_when(
      var == 'chla' ~ 'Chl-a (ug/L)', 
      var == 'tn' ~ 'TN (mg/L)', 
      var == 'tp' ~ 'TP (mg/L)', 
      var == 'nh34' ~ 'NH3, NH4+ (mg/L)',
      var == 'no23' ~ 'Nitrate, nitrite (mg/L)', 
      T ~ var
    )
  )

thm <- theme_minimal() +
  theme(
    legend.title = element_blank(), 
    axis.title = element_blank(), 
    panel.grid.minor.y = element_blank(), 
    strip.placement = 'outside', 
    legend.position = 'top'
  )

p <- ggplot(toplo, aes(x = mo, y = medv, ymin = lov, ymax = hiv, color = typ)) + 
  geom_point(position = position_dodge(width = 0.4), size = 3) + 
  geom_errorbar(width = 0, position = position_dodge(width = 0.4)) + 
  scale_colour_manual(values = c('red', 'black')) + 
  facet_wrap(~lbs, ncol = 1, strip.position = 'left', scales = 'free_y') +
  labs(caption = 'Points are median values, lines are 5th/95th percentiles') +
  thm
p
```

