---
title: Piney Point trends synthesis
author: "MW Beck"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document
---

```{r setup, echo = FALSE, message = F, warning = F, results = 'hide'}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.height = 6, fig.width = 9)

library(tidyverse)
library(lubridate)
library(sf)
library(ggmap)
library(ggspatial)
library(patchwork)
library(units)
library(grid)
library(scales)

data(rswqdat)
data(bswqdat)
data(bsstatloc)
data(rsstatloc)
data(ppseg)
data(parms)
data(rsphydat)
data(rsphypts)
data(trnpts)
data(rstrndat)
data(rstrnpts)
data(bstransect)
data(rstrnwts)

source("R/funcs.R")

# segments
ppseg <- ppseg %>% 
  rename(area = Name) %>% 
  group_by(area) %>% 
  summarise() %>% 
  st_buffer(dist = set_units(0.0001, degree)) %>% 
  st_buffer(dist = set_units(-0.0001, degree)) %>% 
  mutate(
    area = factor(area)
  )

cols <- c("#E16A86", "#50A315", "#009ADE")
names(cols) <- levels(ppseg$area)

# water quality plot fun
wqplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE){

  nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')

  ##
  # wq data
  
  # monitoring data
  rswqtmp <- rswqdat %>% 
    filter(var == vr) %>% 
    filter(!station %in% nonbay) %>% 
    inner_join(rsstatloc, ., by = c('station', 'source')) %>% 
    st_intersection(ppseg) %>% 
    st_set_geometry(NULL) %>% 
    select(-qual, -bswqstation, -nrmrng, -source, -source_lng, -uni, -lbunis) %>% 
    mutate(
      date = floor_date(date, unit = 'week'), 
      mo = month(date), 
      fillcl = factor(area, levels = levels(area), labels = cols), 
      fillcl = as.character(fillcl)
      ) 
    
  # baseline data
  bswqtmp <- bswqdat %>% 
    select(-source, -uni) %>% 
    filter(var == vr) %>% 
    filter(yr > 2005) %>% 
    inner_join(bsstatloc, ., by = 'station') %>% 
    st_intersection(ppseg) %>% 
    st_set_geometry(NULL) %>% 
    group_by(mo, var, area) %>% 
    summarise(   
      avev = mean(val, na.rm = T), 
      stdv = sd(val, na.rm = T), 
      .groups = 'drop'
    ) %>%
    left_join(parms, by = 'var') %>% 
    mutate(
      avev = round(avev, sigdig), 
      stdv = round(stdv, sigdig), 
      minv = avev - stdv, 
      minv = pmax(0, minv),
      maxv = avev + stdv,
      lbunis = gsub('^.*\\s(\\(.*\\))$', '\\1', lbs), 
      lbunis = gsub('pH', '', lbunis), 
      datestr= paste0('2021-', mo, '-01'), 
      datestr = ymd(datestr), 
      dateend = ceiling_date(datestr, unit = 'month')
    )
  
  # boxplot colors
  bxcls <- rswqtmp %>% 
    select(area, date, fillcl) %>% 
    unique
  
  ylab <- unique(rswqtmp$lbs)

  p1 <- ggplot() + 
    geom_rect(data = bswqtmp, aes(xmin = datestr, xmax = dateend, ymin = minv, ymax = maxv, group = mo, fill = 'Monthly baseline (mean +/- 1 sd'), alpha = 0.2) +
    geom_boxplot(data = rswqtmp, aes(x = date, y = val, group = date), fill= bxcls$fillcl, outlier.colour = NA, lwd = 0.75, alpha = 0.8, show.legend = F) + 
    geom_jitter(data = rswqtmp, aes(x = date, y = val, group = date), alpha = 0.4, size = 0.75) + 
    scale_fill_manual(NULL, values = 'blue') +
    scale_linetype_manual(values = 'dashed') + 
    facet_grid(area ~ ., scales = 'free_y') + 
    scale_x_date(breaks = unique(rswqtmp$date), date_labels = '%b %d', expand = c(0.1, 0.1)) +
    labs(
      y = ylab, 
      x = 'Week of'
      ) + 
    coord_cartesian(xlim = range(rswqtmp$date)) +
    theme_minimal(base_size = 14) + 
    theme(
      legend.position = 'top', 
      strip.background = element_blank(), 
      strip.text = element_text(size = 14), 
      axis.text.x = element_text(size = 10, angle = 45, hjust = 1)
    )
  
  if(logtr)
    p1 <- p1 + 
      scale_y_log10(paste0('log-', ylab))
  
 out <- p1
  
 return(out)
 
}

# phyto plot fun
phyplo_fun <- function(rsphydat, rsphypts, ppseg, vr, cols){
    
  dts <- tibble(date = seq.Date(min(rsphydat$date), max(rsphydat$date), by = 'days'))
  
  # monitoring data
  toplo <- rsphydat %>% 
    inner_join(rsphypts, ., by = c('station')) %>% 
    st_intersection(ppseg) %>% 
    st_set_geometry(NULL) %>% 
    select(date, area, species) %>% 
    unique %>% 
    mutate(obs = 1) %>% 
    complete(species, date, area, fill = list(obs = 0)) %>% 
    filter(species == vr) %>% 
    arrange(area, date) %>%
    group_by(area) %>% 
    mutate(
      obs = cumsum(obs)
    )

  p1 <- ggplot(toplo, aes(x = date, y = obs, fill = area)) + 
    geom_area(alpha = 0.8) +
    scale_fill_manual(NULL, values = cols, drop = F) +
    scale_x_date(expand = c(0, 0)) + 
    theme_minimal(base_size = 14) + 
    theme(
      legend.position = 'top', 
      strip.background = element_blank(), 
      strip.text = element_text(size = 14), 
      axis.title.x = element_blank()
    ) + 
    labs(
      y = 'Cumulative observations', 
      caption = 'Observations are based on whether any station in the area recorded the taxon'
    )
  
  return(p1)
  
}
```

# {.tabset}

## Region delineation 

```{r, figh.height = 4, fig.width = 4, fig.margin = T}
##
# map

buffdist <- 0.01
northloc <- 'tr' 
scaleloc <- 'bl'

# layer extent as bbox plus buffer
dat_ext <- ppseg %>% 
  st_bbox %>% 
  st_as_sfc %>% 
  st_buffer(dist = set_units(buffdist, degree)) %>%
  st_bbox %>% 
  unname

# reference data for ggsn, MUST have geometry named column
ggsnref <- ppseg %>% 
  st_bbox %>% 
  st_as_sfc %>%
  st_buffer(dist = set_units(buffdist / 2, degree)) %>% 
  st_as_sf %>%
  st_cast('POINT') %>% 
  rename(geometry = x)

# stamen base map
bsmap1 <- get_stamenmap(bbox = dat_ext, maptype = 'toner-background', zoom = 10)

# change opacity of basemap
mapatt <- attributes(bsmap1)
bsmap1_transparent <- matrix(adjustcolor(bsmap1, 
                                         alpha.f = 0.2), 
                             nrow = nrow(bsmap1))
attributes(bsmap1_transparent) <- mapatt

# plot
map <- ggmap(bsmap1_transparent) +
  geom_sf(data = ppseg, aes(fill = area), color = 'black', inherit.aes = F, alpha = 0.8) +
  geom_sf_text(data = ppseg, aes(label = area), color = 'black', inherit.aes = F, alpha = 0.8, size = 6) +
  scale_fill_manual(values = cols, drop = F) +
  theme(
    legend.title = element_blank(), 
    panel.grid = element_blank(), 
    axis.title = element_blank(), 
    legend.justification = 'top',
    axis.text.y = element_text(size = 8), 
    axis.text.x = element_text(size = 8, angle = 30, hjust = 1),
    panel.background = element_rect(fill = 'white'),
    axis.ticks = element_line(colour = 'grey'),
    panel.border = element_rect(colour = 'grey', fill = NA), 
    legend.position = 'none'
  ) +
  annotation_scale(location = scaleloc) +
  annotation_north_arrow(location = northloc, which_north = "true", height = grid::unit(0.75, "cm"), 
                         width = grid::unit(0.75, "cm"))

map
```

## Water Quality {.tabset .tabset-pills}

### TN

```{r}
wqplo_fun(rswqdat, bswqdat, ppseg, vr = 'tn', cols, logtr = TRUE)
```

### NH3

```{r}
wqplo_fun(rswqdat, bswqdat, ppseg, vr = 'nh34', cols, logtr = TRUE)
```

### TP

```{r}
wqplo_fun(rswqdat, bswqdat, ppseg, vr = 'tp', cols, logtr = TRUE)
```

### Chl

```{r}
wqplo_fun(rswqdat, bswqdat, ppseg, vr = 'chla', cols, logtr = TRUE)
```

### DO Sat.

```{r}
wqplo_fun(rswqdat, bswqdat, ppseg, vr = 'dosat', cols, logtr = FALSE)
```

## Phytoplankton {.tabset .tabset-pills}

### Mixed algae

```{r}
phyplo_fun(rsphydat, rsphypts, ppseg, vr = 'mixed algae', cols)
```

### Centric diatoms

```{r}
phyplo_fun(rsphydat, rsphypts, ppseg, vr = 'Centric Diatoms', cols)
```

### Bacillariophyta

```{r}
phyplo_fun(rsphydat, rsphypts, ppseg, vr = 'Bacillariophyta', cols)
```

### Karenia brevis

```{r}
phyplo_fun(rsphydat, rsphypts, ppseg, vr = 'Karenia brevis', cols)
```

### Pseudo-nitzschia sp.

```{r}
phyplo_fun(rsphydat, rsphypts, ppseg, vr = 'Pseudo-nitzschia sp.', cols)
```

### Rhizosolenia setigera

```{r}
phyplo_fun(rsphydat, rsphypts, ppseg, vr = 'Rhizosolenia setigera', cols)
```

## Seagrass {.tabset .tabset-pills}

### Overall

```{r}
mcrsel <- c("Red", "Green", "Brown", "Cyanobacteria")
savsel <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme')

colpal <- colorRampPalette(RColorBrewer::brewer.pal(n = 8, name = 'Dark2'))
savlevs <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme', 'Ruppia maritima', 'Halophila engelmannii', 'Halophila decipiens')
savcol <- colpal(length(savlevs))
names(savcol) <- savlevs
savcol <- savcol[savsel]
mcrcol <- c('tomato1', 'lightgreen', 'burlywood3', 'lightblue')
names(mcrcol) <- mcrsel
mcrcol <- mcrcol[mcrsel]
cols <- c(mcrcol, savcol)

# add area
trnsum <- rstrndat %>%
  inner_join(rstrnpts, ., by = 'station') %>% 
  st_intersection(ppseg) %>% 
  st_set_geometry(NULL) %>%
  dplyr::group_by(area, typ, date, station, taxa, location) %>%
  dplyr::summarise(
    pa = as.numeric(any(bb > 0))
  ) %>%  
  mutate(
    date = floor_date(date, unit = 'month'), 
    typ = factor(typ, levels = c('mcr', 'sav'), labels = c('Macroalgae', 'Seagrass'))
  ) %>% 
  group_by(area, typ, date, taxa) %>% 
  summarize(
    foest = sum(pa) / length(pa)
  ) %>% 
  filter(taxa %in% c(mcrsel, savsel))

ggplot(trnsum, aes(x =date, y = foest, fill = taxa)) + 
  geom_bar(pch = 16, stat = 'identity', color = 'grey') +
  facet_grid(typ ~ area, scales = 'free') +
  theme_minimal(base_size = 14) + 
  scale_fill_manual(values = cols) +
  labs(
    y = 'Freq. occurrence'
  ) +
  coord_cartesian(ylim = c(0, NA)) +
  theme(
    legend.position = 'top', 
    legend.title = element_blank(),
    strip.background = element_blank(), 
    strip.text = element_text(size = 14), 
    axis.title.x = element_blank(), 
    axis.ticks.x = element_line()
  )
```

### SAV vs macroalgae, by station

```{r}
mcrsel <- c("Red", "Green", "Brown", "Cyanobacteria")
savsel <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme')

# add area
savsum <- rstrndat %>%
  filter(typ == 'sav') %>% 
  inner_join(rstrnpts, ., by = 'station') %>% 
  st_intersection(ppseg) %>% 
  st_set_geometry(NULL) %>% 
  mutate(
    date = floor_date(date, unit = 'week')
  ) %>% 
  dplyr::group_by(area, date, station, taxa) %>%
  dplyr::summarise(
    nsites = length(unique(location)),
    bbest = sum(bb, na.rm = T) / nsites
  ) %>% 
  filter(taxa %in% savsel) %>% 
  group_by(area, date, station) %>% 
  summarize(
    savbbest = mean(bbest)
  )

mcrsum <- rstrndat %>%
  filter(typ == 'mcr') %>% 
  inner_join(rstrnpts, ., by = 'station') %>% 
  st_intersection(ppseg) %>% 
  st_set_geometry(NULL) %>% 
  mutate(
    date = floor_date(date, unit = 'week')
  ) %>% 
  dplyr::group_by(area, date, station, taxa) %>%
  dplyr::summarise(
    nsites = length(unique(location)),
    bbest = sum(bb, na.rm = T) / nsites
  ) %>% 
  filter(taxa %in% mcrsel) %>% 
  group_by(area, date, station) %>% 
  summarize(
    mcrbbest = mean(bbest)
  )

toplo <- full_join(mcrsum, savsum, by = c('area', 'station', 'date')) %>% 
  mutate(
    mo = month(date,label = T)
  )

ggplot(toplo, aes(x = mcrbbest, y = savbbest)) + 
  geom_point(aes(fill = mo), colour = 'grey', pch = 21, size = 3) +
  facet_grid(~ area, scales = 'free') +
  theme_minimal(base_size = 14) + 
  geom_smooth() +
  coord_cartesian(ylim = c(0, NA)) +
  labs(
    y = 'SAV mean abundance (bb)',
    x = 'Macroalgae mean abundance (bb)', 
    caption = 'Results retained by transect'
  ) +
  theme(
    legend.position = 'top', 
    legend.title = element_blank(),
    strip.background = element_blank(), 
    strip.text = element_text(size = 14)
  )
```

### Water quality vs macroalgae vs SAV

```{r}
nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')

# add area
savsum <- rstrndat %>%
  filter(typ == 'sav') %>%  
  inner_join(rstrnpts, ., by = 'station') %>% 
  st_intersection(ppseg) %>% 
  st_set_geometry(NULL) %>% 
  # mutate(
  #   date = floor_date(date, unit = 'week')
  # ) %>% 
  dplyr::group_by(area, date, station, taxa) %>%
  dplyr::summarise(
    nsites = length(unique(location)),
    bbest = sum(bb, na.rm = T) / nsites
  ) %>% 
  filter(taxa %in% savsel) %>% 
  dplyr::group_by(date, area) %>%
  dplyr::summarise(
    savbbest = mean(bbest, na.rm = T),
    .groups = 'drop'
  ) 

mcrsum <- rstrndat %>%
  filter(typ == 'mcr') %>% 
  inner_join(rstrnpts, ., by = 'station') %>% 
  st_intersection(ppseg) %>% 
  st_set_geometry(NULL) %>% 
  # mutate(
  #   date = floor_date(date, unit = 'week')
  # ) %>% 
  dplyr::group_by(area, date, station, taxa) %>%
  dplyr::summarise(
    nsites = length(unique(location)),
    bbest = sum(bb, na.rm = T) / nsites
  ) %>% 
  filter(taxa %in% mcrsel) %>% 
  dplyr::group_by(date, area) %>%
  dplyr::summarise(
    mcrbbest = mean(bbest, na.rm = T),
    .groups = 'drop'
  ) 

rswqtmp <- rswqdat %>% 
  filter(var %in%  c('tn', 'nh34', 'chla', 'tp')) %>% 
  filter(!station %in% nonbay) %>% 
  inner_join(rsstatloc, ., by = c('station', 'source')) %>% 
  st_intersection(ppseg) %>% 
  st_set_geometry(NULL) %>% 
  select(-qual, -bswqstation, -nrmrng, -source, -source_lng, -uni, -lbunis) %>% 
  # mutate(
  #   date = floor_date(date, unit = 'week')
  # ) %>% 
  group_by(date, area, var) %>% 
  summarise(
    val = mean(log(val)), 
    val = exp(val)
    ) %>% 
  spread(var, val)

toplo <- full_join(mcrsum, savsum, by = c('area', 'date')) %>% 
  inner_join(rswqtmp, by = c('area', 'date')) %>% 
  mutate(
    mo = month(date,label = T)
  )

bssz <- 14

mod1 <- summary(lm(chla ~ tn, toplo))
txt <- paste0('Slope ', round(coefficients(mod1)[2, 1], 2), ', R2 ', round(mod1$r.squared, 2), ', pval ', round(coefficients(mod1)[2, 4], 3))
p1 <- ggplot(toplo, aes(x = tn, y = chla)) + 
  geom_point(aes(fill = mo), colour = 'grey', size = 3, pch = 21) + 
  geom_smooth(method = 'lm') + 
  labs(
    x = 'Area averaged TN (mg/L)', 
    y = 'Area averaged Chla (ug/L)', 
    subtitle = txt, 
    title = 'Chla ~ TN'
  ) +
  theme_minimal(base_family = bssz)

mod2 <- summary(lm(mcrbbest ~ tn, toplo))
txt <- paste0('Slope ', round(coefficients(mod2)[2, 1], 2), ', R2 ', round(mod2$r.squared, 2), ', pval ', round(coefficients(mod2)[2, 4], 3))
p2 <- ggplot(toplo, aes(x = tn, y = mcrbbest)) + 
  geom_point(aes(fill = mo), colour = 'grey', size = 3, pch = 21) + 
  geom_smooth(method = 'lm') + 
  labs(
    x = 'Area averaged TN (mg/L)', 
    y = 'Area averaged macroalgae abundance', 
    subtitle = txt, 
    title = 'Macroalgae ~ TN'
  ) +
  theme_minimal(base_family = bssz)

mod3 <- summary(lm(savbbest ~ mcrbbest, toplo))
txt <- paste0('Slope ', round(coefficients(mod3)[2, 1], 2), ', R2 ', round(mod3$r.squared, 2), ', pval ', round(coefficients(mod3)[2, 4], 3))
p3 <- ggplot(toplo, aes(x = mcrbbest, y = savbbest)) +
  geom_point(aes(fill = mo), colour = 'grey', size = 3, pch = 21) +
  geom_smooth(method = 'lm') +
  labs(
    x = 'Area averaged macroalgae abundance',
    y = 'Area averaged SAV abundance',
    subtitle = txt,
    title = 'SAV ~ Macroalgae'
  ) +
  theme_minimal(base_family = bssz)


# mod4 <- summary(lm(chla ~ nh34, toplo))
# txt <- paste0('Slope ', round(coefficients(mod4)[2, 1], 2), ', R2 ', round(mod4$r.squared, 2), ', pval ', round(coefficients(mod1)[2, 4], 3))
# p4 <- ggplot(toplo, aes(x = nh34, y = chla)) + 
#   geom_point(aes(fill = mo), colour = 'grey', size = 3, pch = 21) + 
#   geom_smooth(method = 'lm') + 
#   labs(
#     x = 'Area averaged NH3 (mg/L)', 
#     y = 'Area averaged Chla (ug/L)', 
#     subtitle = txt, 
#     title = 'Chla ~ NH3'
#   ) +
#   theme_minimal(base_family = bssz)
# 
# mod5 <- summary(lm(mcrbbest ~ nh34, toplo))
# txt <- paste0('Slope ', round(coefficients(mod5)[2, 1], 2), ', R2 ', round(mod5$r.squared, 2), ', pval ', round(coefficients(mod5)[2, 4], 3))
# p5 <- ggplot(toplo, aes(x = nh34, y = mcrbbest)) + 
#   geom_point(aes(fill = mo), colour = 'grey', size = 3, pch = 21) + 
#   geom_smooth(method = 'lm') + 
#   labs(
#     x = 'Area averaged NH3 (mg/L)', 
#     y = 'Area averaged macroalgae abundance', 
#     subtitle = txt, 
#     title = 'Macroalgae ~ NH3'
#   ) +
#   theme_minimal(base_family = bssz)
# 
# mod6 <- summary(lm(savbbest ~ mcrbbest, toplo))
# txt <- paste0('Slope ', round(coefficients(mod6)[2, 1], 2), ', R2 ', round(mod6$r.squared, 2), ', pval ', round(coefficients(mod6)[2, 4], 3))
# p6 <- ggplot(toplo, aes(x = mcrbbest, y = savbbest)) + 
#   geom_point(aes(fill = mo), colour = 'grey', size = 3, pch = 21) + 
#   geom_smooth(method = 'lm') + 
#   labs(
#     x = 'Area averaged macroalgae abundance', 
#     y = 'Area averaged SAV abundance', 
#     subtitle = txt, 
#     title = 'SAV ~ Macroalgae'
#   ) +
#   theme_minimal(base_family = bssz)  

p1 + p2 + p3 + plot_layout(ncol = 3, guides = 'collect') & 
  theme(
    legend.position = 'top', 
    legend.title = element_blank()
  )
```

### Water quality vs macroalge vs SAV, correlations

This correlation matrix shows correlations between seagrasses, macroalgae, and water quality.  Seagrass and macroalgae are correlated at the transect scale, i.e., observations are compared at the same location over time.  The water quality comparisons to the other variables are at an aggregated spatial scale because of different sample locations for the transects and water quality.  

```{r, fig.height = 6, fig.width = 6}

mcrsel <- c("Red", "Green", "Brown", "Cyanobacteria")
savsel <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme')
wqsel <- c('chla', 'nh34', 'tn', 'tp')
wqlab <- c('Chl-a', 'NH3, NH4+', 'TN', 'TP')

# correlate transect data at transect scale -------------------------------

# add area
trnsum <- rstrndat %>%
  mutate(
    date = floor_date(date, unit = 'week')
  ) %>%
  dplyr::group_by(date, station, location, taxa) %>%
  dplyr::summarise(
    pa = as.numeric(any(bb > 0))
  ) %>%
  group_by(date, station, taxa) %>% 
  summarize(
    foest = sum(pa) / length(pa)
  )

trncors <- trnsum %>% 
  ungroup %>% 
  filter(taxa %in% c(savsel, mcrsel)) %>% 
  spread(taxa, foest) %>% 
  # filter(area %in% 'Area 3') %>% 
  select(-station, -date)


trncrs <- crossing(var1 = names(trncors), var2 = names(trncors)) %>% 
  filter(var1 != var2) %>% 
  rownames_to_column() %>% 
  group_by(rowname) %>% 
  nest %>% 
  mutate(
    crs = map(data, function(x){
      
      # variables
      vr1 <- trncors[[x$var1]]
      vr2 <- trncors[[x$var2]]
      
      # pearson
      pr_ts <- cor.test(vr1, vr2, method = 'pearson')
      pr_cr <- round(pr_ts$estimate, 2)
      pr_pv <- p_ast(pr_ts$p.value)
      pr <- paste(pr_cr, pr_pv)
      
      out <- data.frame(pr = pr, stringsAsFactors = F)
      return(out)
      
    })
  ) %>% 
  unnest(c('data', 'crs')) %>% 
  ungroup %>% 
  select(-rowname)


# correlate water quality with transect data at aggregated scale ----------

# add area
trnsum <- rstrndat %>% 
  mutate(
    date = floor_date(date, unit = 'week')
  ) %>% 
  inner_join(rstrnpts, ., by = 'station') %>% 
  st_intersection(ppseg) %>% 
  st_set_geometry(NULL) %>% 
  dplyr::group_by(area, date, station, location, taxa) %>%
  dplyr::summarise(
    pa = as.numeric(any(bb > 0))
  ) %>%
  group_by(area, date, taxa) %>% 
  summarize(
    val = sum(pa) / length(pa) # freq occ.
  ) %>% 
  ungroup %>% 
  filter(taxa %in% c(savsel, mcrsel)) %>% 
  rename(var = taxa) %>%
  spread(var, val)

# water quality summary
rswqsum <- rswqdat %>% 
  filter(var %in%  c('tn', 'nh34', 'chla', 'tp')) %>% 
  inner_join(rsstatloc, ., by = c('source', 'station')) %>% 
  st_intersection(ppseg) %>% 
  st_set_geometry(NULL) %>% 
  filter(!station %in% nonbay) %>% 
  select(-qual, -bswqstation, -nrmrng, -source, -source_lng, -uni, -lbunis, -inrng, -lbs) %>% 
  mutate(
    date = floor_date(date, unit = 'week')
  ) %>%
  group_by(area, date, var) %>% 
  summarise(
    val = mean(log(val)), 
    val = exp(val)
  ) %>% 
  spread(var, val)

tocor <- full_join(trnsum, rswqsum, by = c('area', 'date'))

crs <- crossing(var1 = c(wqsel, mcrsel, savsel), var2 = c(wqsel, mcrsel, savsel)) %>% 
  filter(var1 != var2) %>% 
  rownames_to_column() %>% 
  group_by(rowname) %>% 
  nest %>% 
  mutate(
    crs = map(data, function(x){
      
      # variables
      vr1 <- tocor[[x$var1]]
      vr2 <- tocor[[x$var2]]
      
      # pearson
      pr_ts <- cor.test(vr1, vr2, method = 'pearson')
      pr_cr <- round(pr_ts$estimate, 2)
      pr_pv <- p_ast(pr_ts$p.value)
      pr <- paste(pr_cr, pr_pv)
      
      out <- data.frame(pr = pr, stringsAsFactors = F)
      return(out)
      
    })
  ) %>% 
  unnest(c('data', 'crs')) %>% 
  ungroup %>% 
  select(-rowname) %>% 
  filter(!(var1 %in% c(mcrsel, savsel) & var2 %in% c(mcrsel, savsel)))


# combine correlations and make plot --------------------------------------

prplo <- bind_rows(crs, trncrs) %>% 
  separate(pr, c('cor', 'sig'), sep = ' ') %>%  
  mutate(
    cor = as.numeric(cor), 
    var1 = factor(var1, levels = c(wqsel, mcrsel, savsel), labels =  c(wqlab, mcrsel, savsel)), 
    var2 = factor(var2, levels = c(wqsel, mcrsel, savsel), labels =  c(wqlab, mcrsel, savsel)), 
    sig = gsub('ns', '', sig)
  )

pbase <- theme(
  panel.grid.major = element_blank(), 
  panel.grid.minor = element_blank(), 
  axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 8), 
  axis.text.y = element_text(size = 8),
  legend.position = c(0.5, 1.12),
  legend.direction = 'horizontal',
  plot.margin = unit(c(4,4,0,0), "lines"),
  strip.background = element_blank(), 
  strip.text.y = element_text(angle = 0, hjust = 0, vjust = 0.5), 
  panel.background = element_rect(fill = 'black')
) 

outlab <- data.frame(
  y = c(2.5, 6.5, 10), 
  lab = c('Water Quality', 'Macroalgae', 'Seagrasses')
)

p <- ggplot(prplo) + 
  geom_tile(aes(y = var1, x = var2, fill = cor), colour = 'black') + 
  geom_text(aes(y = var1, x = var2, label = sig)) +
  annotation_custom(grob = textGrob(label = outlab$lab[1], hjust = 0, gp = gpar(cex = 0.7)),
                    ymin = outlab$y[1], ymax = outlab$y[1], xmin = 12, xmax = 12) +
  annotation_custom(grob = textGrob(label = outlab$lab[2], hjust = 0, gp = gpar(cex = 0.7)),
                    ymin = outlab$y[2], ymax = outlab$y[2], xmin = 12, xmax = 12) +  
  annotation_custom(grob = textGrob(label = outlab$lab[3], hjust = 0, gp = gpar(cex = 0.7)),
                    ymin = outlab$y[3], ymax = outlab$y[3], xmin = 12, xmax = 12) +
  annotation_custom(grob = textGrob(label = outlab$lab[1], hjust = 0.5, gp = gpar(cex = 0.7)),
                    xmin = outlab$y[1], xmax = outlab$y[1], ymin = 12, ymax = 12) +
  annotation_custom(grob = textGrob(label = outlab$lab[2], hjust = 0.5, gp = gpar(cex = 0.7)),
                    xmin = outlab$y[2], xmax = outlab$y[2], ymin = 12, ymax = 12) +  
  annotation_custom(grob = textGrob(label = outlab$lab[3], hjust = 0.5, gp = gpar(cex = 0.7)),
                    xmin = outlab$y[3], xmax = outlab$y[3], ymin = 12, ymax = 12) +
  pbase +
  scale_y_discrete('', expand = c(0, 0)) + #, labels = parse(text = rev(labs))) + 
  scale_x_discrete('', expand = c(0, 0)) + #, labels = parse(text = rev(labs))) +
  scale_fill_gradientn('Correlation', colours = c(muted("blue"), "white", muted("red")), limits = c(-1, 1)) +
  guides(fill = guide_colourbar(barheight = 0.5, barwidth = 5, label.theme = element_text(size = 6, angle = 0))) +
  geom_hline(yintercept = 4.5, size = 1.5) +
  geom_hline(yintercept = 8.5, size = 1.5) +
  geom_vline(xintercept = 4.5, size = 1.5) +
  geom_vline(xintercept = 8.5, size = 1.5)

# Code to override clipping
gt <- ggplot_gtable(ggplot_build(p))
gt$layout$clip[gt$layout$name == "panel"] <- "off"
grid.draw(gt)
```

### Biomass estimates {#biomass}

Simple regressions of weight vs abundance by major macroalgae group.  Regressions are fit through the origin so that zero weight is fixed at zero mass.

```{r, fig.height = 5, fig.width = 4}
rssub <- rstrndat %>% 
  filter(typ == 'mcr') %>% 
  select(station, location, date, taxa, bb)

bssub <- bstransect %>% 
  filter(var == 'Abundance') %>% 
  select(
    station = Transect, 
    location = Site, 
    date = Date,
    taxa = Savspecies,
    bb = aveval
  ) %>% 
  filter(date %in% rstrnwts$date & station %in% rstrnwts$station & location %in% rstrnwts$location) %>% 
  filter(grepl('^DA', taxa)) %>% 
  mutate(
    location = as.numeric(location), 
    taxa = case_when(
      grepl('Red', taxa) ~ 'Red', 
      grepl('Green', taxa) ~ 'Green', 
      grepl('Macroalgae', taxa) ~ 'Red',
      T ~ taxa# verified that the location/date/transect match with weights was red, only one instance
    )
  ) 
  
tojn <- bind_rows(rssub, bssub)

wtssub <- rstrnwts %>% 
  select(-genus) %>% 
  rename(taxa = group) %>% 
  filter(!grepl('and', taxa)) %>% 
  left_join(tojn, by = c('station', 'date', 'location', 'taxa')) %>% 
  mutate(
    weight = weight_g * 0.004 # g / 0.25m2 to kg / m2
  )

ymax <- wtssub %>% 
  pull(weight) %>% 
  max(na.rm = T)

ggplot(wtssub, aes(x = bb, y = weight)) + 
  geom_point() +
  facet_wrap(~taxa, ncol = 1) +
  stat_smooth_func_with_pval(geom = "text", method = "lm", hjust = 0, parse = T, xpos = 0, xpos2 = 0, ypos = ymax, ypos2 = 0.8 * ymax, formula = y ~ 0 + x) +
  geom_smooth(method = 'lm', se = F, formula = y ~ 0 + x) + 
  labs(
    x = 'Abundance (bb)', 
    y = 'Weight (kg / m2)'
  ) + 
  theme_minimal() + 
  theme(
    strip.background = element_blank(), 
    panel.grid.minor = element_blank()
  )
```

Using the regression, a weight was assigned to all macroalgae samples where abundance (bb) was recorded.  The weights were then summed across all quadrats by transect, macroalgae group, month, and region ([here](https://tbep-tech.github.io/piney-point-analysis/trends#region-delineation)) to represent total sampled biomass. However, this is biased by relative sample effort (Area 1 has been sampled ~3x more than Area 3).  Below, the weights are standardized to unit effort.

```{r, fig.height = 3, fig.width = 5}
wtsmod <- wtssub %>% 
  group_by(taxa) %>% 
  nest() %>% 
  mutate(
    mod = purrr::map(data, lm, formula = weight ~ 0 + bb)
  ) %>% 
  select(taxa, mod)

wtsest <- rstrndat %>% 
  filter(taxa %in% c('Red', 'Green', 'Cyanobacteria')) %>% 
  group_by(taxa) %>% 
  nest %>% 
  left_join(wtsmod, by = 'taxa') %>% 
  mutate(
    weight_kgm2 = purrr::pmap(list(object = mod, newdata = data), predict)
  ) %>% 
  select(-mod) %>% 
  unnest(c('data', 'weight_kgm2'))
    
mcrsel <- c("Red", "Green", "Brown", "Cyanobacteria")
mcrcol <- c('tomato1', 'lightgreen', 'burlywood3', 'lightblue')
names(mcrcol) <- mcrsel
mcrcol <- mcrcol[unique(wtsest$taxa)]

# add area
wtsarea <- wtsest %>%
  inner_join(rstrnpts, ., by = 'station') %>% 
  st_intersection(ppseg) %>% 
  st_set_geometry(NULL) %>%
  dplyr::group_by(area, typ, date, station, taxa, location) %>%
  mutate(
    dateflr = floor_date(date, unit = 'month')
  ) 

wtssum <- wtsarea %>% 
  group_by(area, dateflr, taxa) %>% 
  summarize(
    kgest = sum(weight_kgm2) / 4 # divide by four to get total biomass across all quads (0.25 m2)
  )

ggplot(wtssum, aes(x = dateflr, y = kgest, fill = taxa)) + 
  geom_bar(pch = 16, stat = 'identity', color = 'grey') +
  facet_grid(~ area, scales = 'free') +
  theme_minimal(base_size = 14) + 
  scale_fill_manual(values = mcrcol) +
  labs(
    y = 'Total sampled weight (kg)'
  ) +
  coord_cartesian(ylim = c(0, NA)) +
  theme(
    legend.position = 'top', 
    legend.title = element_blank(),
    strip.background = element_blank(), 
    strip.text = element_text(size = 14), 
    axis.title.x = element_blank(), 
    axis.ticks.x = element_line()
  )
```

Because of different sampling effort by month and location, weights were standardized by dividing by sample effort (number of unique transect visits in a month).  This can also be considered as kg / 1.25 m2 because five quadrats of 0.25 m2 are surveyed each transect.  So converting this to kg / m2, the weights are multiplied by 0.8 (1 / 1.25).

```{r, fig.height = 3, fig.width = 5}
wtssum <- wtsarea %>% 
  unite('datetrn', date, station, remove = F) %>% 
  group_by(area, dateflr, taxa) %>% 
  summarize(
    kgest = sum(weight_kgm2) / 4 / length(unique(datetrn)), 
    kgest = kgest * 0.8
  )

ggplot(wtssum, aes(x = dateflr, y = kgest, fill = taxa)) + 
  geom_bar(pch = 16, stat = 'identity', color = 'grey') +
  facet_grid(~ area, scales = 'free') +
  theme_minimal(base_size = 14) + 
  scale_fill_manual(values = mcrcol) +
  labs(
    y = 'Biomass (kg / m2)'
  ) +
  coord_cartesian(ylim = c(0, NA)) +
  theme(
    legend.position = 'top', 
    legend.title = element_blank(),
    strip.background = element_blank(), 
    strip.text = element_text(size = 14), 
    axis.title.x = element_blank(), 
    axis.ticks.x = element_line()
  )

```

